A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex

\hypertarget{covid-19-in-spain}{%
\subsection{Covid-19 in Spain}\label{covid-19-in-spain}}

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4805.98 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is an \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in province
\(s\text{ }s=1,...,N\)) on day \(t\text{ }(t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is an \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the usual vector of
random residuals.

A key feature of the SUR framework is the dependence structure among the
vectors of random residuals, which takes the following form:

@@

Note that this specification is very flexible, since it allows the
coefficients \(\beta_{it}\) to vary in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). A less flexible if more parsimonious model can
be obtained by imposing restrictions to some \(\beta\) coefficients to
fix them across equations. This is equivalent to specifying an effect
that is constant in time. In this case, we can reformulate the
coefficients expression of
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant.

Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:

@@

\noindent where\ldots.

As is the case with cross-sectional data, it is possible to test the
residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
several tests have been developed to test the null hypothesis of spatial
independence (see López et al., 2014). When the null hypothesis is
rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. A sp

The stack expresion for the spatial lag SUR model (SUR-SLM),

@@

\noindent where \(A =I_{TN}-\bf{\rho} \otimes W\) with
\(\bf{\rho} = diag(\rho_1, \cdots, \rho_T)\).

@@

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)).

Another alternative methodologies could be use. By example, a dynamic
spatial panel methodology with fixed spatial an temporal effects
(e.g.~Elhorst 2014, Cap. 4), but those models are more dont take account
correlation between errors. Therefore, a spatial SUR approach is more
reasonable for our purpose.

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. It can be seen that the highest
incidence at this early date was in the provice of Álava, in the North
of Spain. While not the most populous province, with a population of
only 331,549, Álava has the highest GDP per capita of all provinces.
Vitoria, its main city, is the capital of the Basque country and has
been the focus of efforts to develop a ``Global Basque City'' (Meijers
et al., 2008), along with San Sebastian and Bilbao. The other early
focus of the pandemic in Spain was Madrid, which is the most populous
province in the country and has the second highest GDP per capita after
Álava. The early outbreaks in these two provinces can be traced
throughout the progression of the pandemic over time, although by the
end of the period under consideration, other provinces had matched and
even surpased their incidence rates, including Segovia and Soria to the
north of Madrid, and Ciudad Real and Albacete to the south.

@@

@@

@@

@@

>>>>

\#\# Covid-19 in Spain

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

A key feature of the SUR model is the dependence structure among the
vectors of residuals, namely:

@@

Note that this specification is very flexible, in that it allows changes
in the coefficients \(\beta_{it}\) in order to modulate the effect of
\(X^i_{.t}\) on \(y_t\). This flexibility can be reduced and it is
posible to impose restrinctions considering some \(\beta\) coefficients
as being constant over time. In this case, we can reformulate the
coefficients expression
\(\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})\)
to restrict the first \(r\) coefficients to be constant across
equations. This is equivalent to specifying some effects to be invariant
over time.

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

@@

As is the case with cross-sectional data, it is possible to test the
residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
several tests have been developed to test the null hypothesis of spatial
independence (see López et al., 2014). When the null hypothesis is
rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. Spatial analytical
approaches were used to understand contagion-difussion processes in the
case of infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

The stack expresion for the spatial lag SUR model (SUR-SLM) is as
follows:

@@

\noindent where \(A =I_{TN}-\bf{\Lambda} \otimes W\) with
\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).

@@

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
alternative methodologies could be use. By example, a dynamic spatial
panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
2014, Cap. 4), but those models are more dont take account correlation
between errors. Therefore, a spatial SUR approach is more reasonable for
our purpose.

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. Our analysis begins on March 13, which
is the first date when every province had reported at least one case of
COVID-19. It can be seen that the highest incidence at this early date
was in the provice of Álava, in the North of Spain. While not the most
populous province, with a population of only 331,549, Álava has the
highest GDP per capita of all provinces. Vitoria, its main city, is the
capital of the Basque country and has been the focus of efforts to
develop a ``Global Basque City'' (Meijers et al., 2008), along with San
Sebastian and Bilbao. The other early focus of the pandemic in Spain was
Madrid, which is the most populous province in the country and has the
second highest GDP per capita after Álava. The early outbreaks in these
two provinces can be traced throughout the progression of the pandemic
over time, although by the end of the period under consideration, other
provinces had matched and even surpased their incidence rates, including
Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
to the south.

@@

\leavevmode\hypertarget{ref-Cao2010spatio}{}%
Cao, Z., Zeng, D., Zheng, X., Wang, Q., Wang, F., Wang, J., Wang, X.,
2010. Spatio-temporal evolution of beijing 2003 sars epidemic. Science
China Earth Sciences 53, 1017--1028.
doi:\href{https://doi.org/10.1007/s11430-010-0043-x}{10.1007/s11430-010-0043-x}

@@

\leavevmode\hypertarget{ref-Cliff1998detecting}{}%
Cliff, A., Haggett, P., Smallman-Raynor, M., 1998. Detecting
space---time patterns in geocoded disease data. Cholera in london, 1854
measles in the united states, 1962--95, in: Geomed'97. Springer, pp.
13--42.

@@

\leavevmode\hypertarget{ref-Desjardins2020rapid}{}%
Desjardins, M., Hohl, A., Delmelle, E., 2020. Rapid surveillance of
covid-19 in the united states using a prospective space-time scan
statistic: Detecting and evaluating emerging clusters. Applied Geography
102202.

@@

\leavevmode\hypertarget{ref-Meng2005understanding}{}%
Meng, B., Wang, J., Liu, J., Wu, J., Zhong, E., 2005. Understanding the
spatial diffusion process of severe acute respiratory syndrome in
beijing. Public Health 119, 1080--1087.
doi:\href{https://doi.org/https://doi.org/10.1016/j.puhe.2005.02.003}{https://doi.org/10.1016/j.puhe.2005.02.003}
